---
title: "Transparent O*NET to NOC 2021 Mapping"
format:
  html:
    code-fold: true
---

```{r, include=FALSE}
library(tidyverse)
library(reactable)
library(fs)
# find latest version directory under output/
version_dirs <- fs::dir_ls("output", type = "directory")

version_tbl <- tibble(
  path = version_dirs,
  version = fs::path_file(path)
) |>
  mutate(valid = grepl("^\\d+\\.\\d+\\.\\d+$", version)) |>
  filter(valid)

if (nrow(version_tbl) == 0) {
  stop("No valid version directories found under 'output/'")
}

versions <- numeric_version(version_tbl$version)

latest_dir <- version_tbl$path[order(versions, decreasing = TRUE)[1]]
latest_version <- fs::path_file(latest_dir)
message("Using data from: ", latest_dir)

onet_to_noc2021_paths <- read_csv(paste0(latest_dir,"/onet_to_noc2021_paths.csv"))
onet_data <- read_csv(paste0(latest_dir,"/onet_data.csv"))
mapping <- read_csv(paste0(latest_dir,"/onet_to_noc2021_mapping.csv"))
mapping_strength <- read_csv(paste0(latest_dir,"/onet_to_noc2021_mapping_strength.csv"))
# -----------------------------------------------------------------
# Naive mapping 
#-----------------------------------------------------------------

naive_mapping <- onet_to_noc2021_paths|>
  select(onet_soc_code, onet_plus_title, noc_2021, noc_plus_title)

# ---------------------------------------------------------------
# Comparison between down_weight and naive_mapping
#----------------------------------------------------------------

naive_skills <- left_join(naive_mapping, onet_data, by = join_by(onet_soc_code)) |>
  select(-onet_soc_code) |>
  group_by(noc_plus_title) |>
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))

downweight_skills <- mapping|>
  select(onet_soc_code, noc_plus_title, down_weight)|>
  left_join(onet_data, by = join_by(onet_soc_code))|>
  mutate(across(where(is.numeric) & !all_of("down_weight"), ~ .x * down_weight)) |>
  group_by(noc_plus_title) |>
  summarise(across(where(is.numeric), ~ sum(.x, na.rm = TRUE)))|>
  select(-down_weight)

#ensure alignment------------------------

naive_mat <- naive_skills |>
  column_to_rownames("noc_plus_title")

downweight_mat <- downweight_skills |>
  column_to_rownames("noc_plus_title")

common_noc <- intersect(rownames(naive_mat), rownames(downweight_mat))
common_vars <- intersect(colnames(naive_mat), colnames(downweight_mat))

naive_mat <- naive_mat[common_noc, common_vars]
downweight_mat <- downweight_mat[common_noc, common_vars]

stopifnot(all(rownames(naive_mat) == rownames(downweight_mat)))

delta_euclid <- sqrt(rowSums((naive_mat - downweight_mat)^2))

tbbl_delta <- tibble(
  noc_plus_title = rownames(naive_mat),
  delta_euclid = delta_euclid
)

tbbl_compare <- tbbl_delta |>
  left_join(mapping_strength, by = join_by(noc_plus_title))

```

## Data version

This document uses mapping version: **`r latest_version`**


## 1. Why not just use OaSIS?

OaSIS provides occupation-level skill profiles for Canadian classifications, making it an attractive alternative. However, it does not expose how those profiles are constructed.

- No observable mapping between O*NET and NOC occupations  
- No weights linking underlying occupations  
- No information on how multiple occupations are combined  

As a result, users cannot audit or replicate how each skill profile is formed.

This project instead constructs the mapping explicitly, allowing:

- full transparency of inputs and weights  
- diagnostic evaluation of mapping quality  
- reproducibility of all outputs  

---

## 2. Why naive crosswalk chaining fails

A common approach is to join crosswalks:

O*NET → SOC → NOC

This produces many-to-many expansions and implicitly assigns equal weight across all paths.

This leads to:

- over-weighting of occupations that appear frequently in crosswalk chains  
- inclusion of unrelated occupations  
- distorted skill vectors 

For instance, consider 72600: Air pilots, flight engineers and flying instructors. The naive 
mapping would apply equal weighting to the following rows:

```{r}
naive_mapping|> 
  filter(noc_plus_title == "72600: Air pilots, flight engineers and flying instructors")|>
  select(onet_plus_title, noc_plus_title)|>
  reactable(
  columns = list(onet_plus_title = colDef(name = "O*NET occupation"),
                 noc_plus_title = colDef(name = "NOC")),
  searchable = FALSE,
  filterable = FALSE,
  sortable = TRUE,
  pagination = FALSE,
  bordered = TRUE,
  striped = TRUE,
  highlight = TRUE
)
```

---

Only $\frac38$ of the skill profile for NOC 72600: Air pilots, flight engineers and flying instructors is coming from pilots! The other $\frac58$ is derived from the skill profiles of agricultural workers.  

---

## 3. Mapping construction

The mapping is constructed in three stages:

1. O*NET 2019 → SOC 2018  
2. SOC 2018 → NOC 2016  
3. NOC 2016 → NOC 2021  

Each stage introduces potential one-to-many relationships.

### Path weights

Weights are assigned as:

$$
w_{path} = w_1 \times w_2 \times w_3
$$

where each stage distributes mass equally across branches.

### Aggregation

Weights are aggregated to the NOC level and normalized.

### Downweighting

A power transformation is applied:

$$
\tilde{w}_k \propto w_k^{2}
$$

to reduce the influence of small-weight links.

Now we can return to our 72600: Air pilots, flight engineers and flying instructors example, and investigate these weights.

```{r}
mapping|> 
  filter(noc_plus_title == "72600: Air pilots, flight engineers and flying instructors")|>
  select(onet_plus_title, noc_plus_title, base_weight, down_weight)|>
  reactable(
  columns = list(onet_plus_title = colDef(name = "O*NET occupation"),
                 noc_plus_title = colDef(name = "NOC"),
                 base_weight= colDef(name = "Base Weight", format = colFormat(digits = 2)),
                 down_weight= colDef(name = "Down Weight", format = colFormat(digits = 2))
                 ),
  searchable = FALSE,
  filterable = FALSE,
  sortable = TRUE,
  pagination = FALSE,
  bordered = TRUE,
  striped = TRUE,
  highlight = TRUE
)
```

---

Using the base weights, 78% of the skill profile is being derived from pilots.  Recall in the naive case, only 37.5% of the skill profile came from pilots. Things improve further if one uses the down weight: now 96% of the skill profile is derived from pilots.  This is the most extreme example of a difference in skill profile between the naive weights and either the base or down weights. 

---

## 4. Outputs

The build process produces:

- `onet_to_noc2021_paths`  
- `mapping`  
- `mapping_strength`  

Each output is fully reproducible from raw inputs.

---

## 5. Diagnostics for interpretation

Two key diagnostics are provided:

### Mapping concentration (Herfindahl)

$$
H = \sum_k w_k^2
$$

Interpretation:

- High H → well-defined occupation  
- Low H → heterogeneous occupation  

### Sensitivity to naive mapping (Δ)

$$
\Delta = || s_{naive} - s_{weighted} ||
$$

Interpretation:

- Large Δ → naive mapping is misleading  
- Small Δ → naive mapping is acceptable  

---

## 6. Interpreting H and Δ

These measure different things:

- **H**: whether a single skill profile is meaningful  
- **Δ**: whether naive mapping introduces error  

### Cases

- High H, low Δ → stable and well-defined  
- High H, high Δ → naive mapping contaminated  
- Low H, NOC is inherently too broad, no way to fix  

### Plot

```{r}
plt <- ggplot(tbbl_compare, aes(x = down_herf, y = delta_euclid, label=noc_plus_title)) +
  geom_point(alpha = 0.25) +
  labs(
    x = "Herfindahl (mapping concentration)",
    y = "Change in skill vector (Euclidean distance)"
  )
plotly::ggplotly(plt)
```



---

## 7. Low H example

41200: University professors and lecturers are a good example of a NOC where there 
is no solution to the problem that this group is just too diverse to be represented by a
single skill profile. 

```{r}
mapping|> 
  filter(noc_plus_title == "41200: University professors and lecturers")|>
  select(onet_plus_title, noc_plus_title, base_weight, down_weight)|>
  reactable(
  columns = list(onet_plus_title = colDef(name = "O*NET occupation"),
                 noc_plus_title = colDef(name = "NOC"),
                 base_weight= colDef(name = "Base Weight", format = colFormat(digits = 2)),
                 down_weight= colDef(name = "Down Weight", format = colFormat(digits = 2))
                 ),
  searchable = FALSE,
  filterable = FALSE,
  sortable = TRUE,
  pagination = TRUE,
  bordered = TRUE,
  striped = TRUE,
  highlight = TRUE
)
```
---

## 8. Guidance for use

- Use down weight mapping as default  
- Treat mapped skill profiles of low-H occupations with caution  
- Avoid naive mapping  

---

## 9. Reproducibility

All results are generated from:

- O*NET data  
- published crosswalks  

The full pipeline is contained in the build script.

Version: `r latest_version`
