---
title: "Transparent O*NET to NOC 2021 Mapping"
format:
  html:
    self-contained: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false    
---

```{r, include=FALSE}
library(tidyverse)
library(reactable)
library(plotly)
library(fs)
library(conflicted)
conflicts_prefer(
  dplyr::filter,
  dplyr::select,
  dplyr::mutate,
  dplyr::summarise,
  dplyr::arrange
)

# find latest version directory under output/
version_dirs <- fs::dir_ls("output", type = "directory")

version_tbl <- tibble(
  path = version_dirs,
  version = fs::path_file(path)
) |>
  mutate(valid = grepl("^\\d+\\.\\d+\\.\\d+$", version)) |>
  filter(valid)

if (nrow(version_tbl) == 0) {
  stop("No valid version directories found under 'output/'")
}

versions <- numeric_version(version_tbl$version)

latest_dir <- version_tbl$path[order(versions, decreasing = TRUE)[1]]
latest_version <- fs::path_file(latest_dir)
message("Using data from: ", latest_dir)

required_files <- c(
  "onet_to_noc2021_mapping.csv",
  "diagnostics.csv",
  "onet_to_noc2021_paths.csv"
)

missing_files <- required_files[!file.exists(file.path(latest_dir, required_files))]

if (length(missing_files) > 0) {
  stop("Missing expected output files: ", paste(missing_files, collapse = ", "))
}

mapping <- read_csv(paste0(latest_dir,"/onet_to_noc2021_mapping.csv"))
diagnostics <- read_csv(paste0(latest_dir,"/diagnostics.csv"))
onet_to_noc2021_paths <- read_csv(paste0(latest_dir, "/onet_to_noc2021_paths.csv"))
```

## Data version

This document uses mapping version: **`r latest_version`**


## 1. Why not just use OaSIS?

OaSIS provides occupation-level skill profiles for Canadian classifications, making it an attractive alternative. However, it does not expose how those profiles are constructed.

- No observable mapping between O*NET and NOC occupations  
- No weights linking underlying occupations  
- No information on how multiple occupations are combined  

As a result, users cannot audit or replicate how each skill profile is formed.

This project instead constructs the mapping explicitly, allowing:

- full transparency of inputs and weights  
- diagnostic evaluation of mapping quality: in particular, whether mapped occupations form a coherent cluster in skill space  
- reproducibility of all outputs  

---

## 2. Why naive crosswalk chaining fails

A common approach is to join crosswalks:

O*NET → SOC → NOC

This produces many-to-many expansions and implicitly assigns equal weight across all admissible paths.

This leads to:

- over-weighting of occupations that appear frequently in crosswalk chains  
- inclusion of unrelated occupations  
- distorted skill vectors 

## 3. Weighted mapping construction

The mapping is constructed in three stages:

1. O*NET 2019 → SOC 2018  
2. SOC 2018 → NOC 2016  
3. NOC 2016 → NOC 2021  

Each stage introduces potential one-to-many relationships.

### Path weights

Weights are assigned as:

$$
w_{path} = w_1 \times w_2 \times w_3
$$

where each stage distributes mass equally across branches.

### Example

Consider 72600: Air pilots, flight engineers and flying instructors.  The table below
features two different weighting notions.  Naive weights put an equal weight on all paths^[Note that there are two separate paths from 53-2012.00: Commercial Pilots.] from O*NET occupations to 72600: Air pilots, flight engineers and flying instructors.   

In contrast, path weights indicate the extent of splitting across the crosswalks: 

- For 53-2011.00: Airline Pilots, Copilots, and Flight Engineers there are no splits along the crosswalk: the only path is to 72600: Air pilots, flight engineers and flying instructors, so this path gets a weight of 1. 
- For the first path leading from 53-2012.00: Commercial Pilots there is a 2 way split, so the path weight = .5. 
- For the second path leading from 53-2012.00: Commercial Pilots there is both a 2 way and a 4 way split, so the path weight is .125.
- Notice that path weights tend to be smaller than the naive weights for the O\*NET agricultural occupations.  


```{r}
onet_to_noc2021_paths|> 
  filter(noc_plus_title == "72600: Air pilots, flight engineers and flying instructors")|>
  select(onet_plus_title, noc_plus_title, equal_weight, path_weight)|>
  janitor::adorn_totals()|>
  reactable(
  columns = list(onet_plus_title = colDef(name = "O*NET occupation"),
                 noc_plus_title = colDef(name = "NOC"),
                 equal_weight =colDef(name = "Naive weights"),
                 path_weight =colDef(name = "Path weights", format = colFormat(digits = 4))
                 ),
  searchable = FALSE,
  filterable = FALSE,
  sortable = TRUE,
  pagination = FALSE,
  bordered = TRUE,
  striped = TRUE,
  highlight = TRUE
)
```

---

If we use the naive weights to create an average skill profile for NOC 72600: Air pilots, flight engineers and flying instructors, it would mostly ($\frac58$) reflect agricultural workers. 

---


### Base weights

- To derive our "Base weights" from the above table, first we aggregate path weights by origin and destination, and then normalize so they sum to one. 

- e.g. from above we see that the total path weight is 2.1042, so 53-2011.00: Airline Pilots, Copilots, and Flight Engineers base weight is $\frac{1}{2.1042}=0.48$.

- Similarly 53-2012.00: Commercial Pilots have an *aggregate* path weight of .625, so its base weight is $\frac{.625}{2.1042}=0.3$. 

- Note that by switching from naive weights to path weights 78% of the skill profile is based on the skills of... pilots.  


### Downweighting

Finally, a power transformation and re-normalization is undertaken to further diminish the weights on mapping artifacts (in this case,  agricultural occupations)

$$
\tilde{w}_k \propto w_k^{2}
$$


The full table of weights for 72600: Air pilots, flight engineers and flying instructors is below:

```{r}
mapping|> 
  filter(noc_plus_title == "72600: Air pilots, flight engineers and flying instructors")|>
  select(onet_plus_title, noc_plus_title, base_weight, down_weight)|>
  reactable(
  columns = list(onet_plus_title = colDef(name = "O*NET occupation"),
                 noc_plus_title = colDef(name = "NOC"),
                 base_weight= colDef(name = "Base Weight", format = colFormat(digits = 2)),
                 down_weight= colDef(name = "Down Weight", format = colFormat(digits = 2))
                 ),
  searchable = FALSE,
  filterable = FALSE,
  sortable = TRUE,
  pagination = FALSE,
  bordered = TRUE,
  striped = TRUE,
  highlight = TRUE
)
```

---

Using the down weights results in 96% of the skill profile being based on pilots.  This example is not typical, but is illustrative of the problem associated with using naive weights. 

---

## 4. Outputs

The build process produces:


- `onet_to_noc2021_mapping.csv`  
- `diagnostics.csv`  
- `onet_to_noc2021_paths.csv`

Each output is fully reproducible from raw inputs.

---

## 5. Diagnostics for interpretation

Two key diagnostics are provided:

### Mapping concentration (Herfindahl)

$$
H = \sum_k w_k^2
$$

Interpretation:

- High H → well-defined occupation  
- Low H → heterogeneous occupation  

### Scaled Similarity

We represent each O\*NET occupation as a point in a reduced skill space (using the first 10 principal components), and for each NOC we compute a weighted centroid of these points using the mapping weights. This centroid can be interpreted as the “average” skill profile implied by the mapping. We then measure how far each contributing O*NET occupation lies from this centroid and take a weighted average of those distances. The resulting statistic captures the coherence of the mapping in skill space: low values indicate that the mapped occupations form a tight, internally consistent cluster, while high values indicate that the NOC is being represented by a set of occupations with very different skill profiles.

As a final transformation we take the percentile value of the above statistic, and subtract it from one, and call it scaled similarity. This puts it on a similar scale to the Herfindahl index.  We sort our diagnostic table on the product of Herfindahl and this scaled similarity.  

---

## 6. Interpreting Herfindahl and Scaled Similarity

These measure different things:

- **Herfindahl**: how diffuse/concentrated is the NOC in terms of O\*NET contributions 
- **Scaled Similarity**: how similar the O\*NET contributions are. 

### Cases

- High Herfindahl, any value of scaled similarity: no problem, almost all weight put on a single O\*NET occupation.
- Low Herfindahl, high scaled similarity: no problem, a diffuse weighted average of highly similar O\*NET occupations. 
- Low Herfindahl, low scaled similarity: big problem, a diffuse weighted average of disparate O\*NET occupations.

### Plot

```{r}
plt <- diagnostics|>
  ggplot(aes(`Herfindahl Index`, `Scaled Similarity`))+
  coord_equal() +
  coord_cartesian(xlim = c(-.1, 1.1), ylim = c(-.1, 1.1)) +  
  geom_jitter(aes(fill=`Share of weights reallocated`, label=noc_plus_title), shape=21, width=.01, height=.01, size=2)+
  scale_fill_viridis_c()+
  labs(title="Mapping Quality: Problematic NOCs close to origin(low concentration, low similarity)")

ggplotly(plt) 
```



---

## 7. Low Herfindahl, low scaled similarity example

41200: University professors and lecturers are a good example of the mapping problem: diffuse weighting of 
disparate skill profiles.

```{r}
mapping|> 
  filter(noc_plus_title == "41200: University professors and lecturers")|>
  select(onet_plus_title, noc_plus_title, base_weight, down_weight)|>
  reactable(
  columns = list(onet_plus_title = colDef(name = "O*NET occupation"),
                 noc_plus_title = colDef(name = "NOC"),
                 base_weight= colDef(name = "Base Weight", format = colFormat(digits = 2)),
                 down_weight= colDef(name = "Down Weight", format = colFormat(digits = 2))
                 ),
  searchable = FALSE,
  filterable = FALSE,
  sortable = TRUE,
  pagination = TRUE,
  bordered = TRUE,
  striped = TRUE,
  highlight = TRUE
)
```
---

## 8. Guidance for use

- Use down weight mapping as default  
- Treat mapped skill profiles of low-H low similarity occupations with caution  
- Avoid naive mapping  

---

## 9. Reproducibility

All results are generated from:

- O*NET data  
- published crosswalks  

The full pipeline is contained in the build script.

Version: `r latest_version`
